\section*{Abstract}
\label{sec:abstract}

Chest radiography is the most common first-line imaging examination, yet most automated chest X-ray (CXR) classifiers are image-centric and ignore the clinical context that radiologists routinely use, including the indication for the examination, vital signs, and prior imaging studies. This thesis studies whether such clinical context and prior-study information improve multi-label CXR classification under a long-tailed label distribution. Building on the CaMCheX framework, we design and evaluate a prior-aware multimodal model that encodes current frontal and lateral radiographs, pre-interpretation clinical indication text, structured emergency-department vital signs, and prior-study context (prior images, reports, labels, and time-gap embeddings) with modality-specific encoders, and fuses them through an asymmetric transformer-decoder in which the current study attends to pooled prior context. Prior dropout is applied during training to limit over-reliance on historical information, and the radiology-domain text encoder is fine-tuned end-to-end in the final model. Experiments are conducted on a filtered CXR-LT-2023-aligned dataset built from MIMIC-CXR, MIMIC-IV, and MIMIC-IV-ED, with 26 findings evaluated overall and by head, medium, and tail frequency groups. The proposed model achieves 0.5662 mean Average Precision (mAP) and 0.9026 macro AUROC, substantially outperforming a reproduced image-only CaMCheX baseline (0.3246 mAP) and a multimodal variant without prior context (0.4098 mAP), while using a smaller architecture than the CaMCheX-style reference (164.58M vs.\ 229.71M total parameters) at half the input resolution. Gains are largest on rare findings: tail-class mAP improves from 0.0827 (image-only) and 0.1586 (no-prior multimodal) to 0.4076. Ablations show that clinical indication text, prior-study integration, and text-encoder fine-tuning drive most of the improvement, and qualitative Grad-CAM and token-level attribution analyses are used as sanity checks of model behaviour. These results indicate that prior-aware multimodal fusion is a practical and effective direction for clinically contextualized, long-tailed chest X-ray classification.