\section{Methodology}
\label{sec:methodology}

In this section, we present the methodology for the clinically-aligned multi-modal chest X-ray classification framework, which we refer to as \textbf{CaMCheX}. We detail the formal problem definition, describe the individual modal encoders (image, text, and vitals), outline the token fusion strategies (comparing the spatially expanded V2 and the collapsed single-token V3 architectures), describe the prior-aware temporal modeling extensions (comparing the concatenated self-attention V3 configuration and the asymmetric cross-attention V4 architecture), and discuss the training objective.

An overview of the system architecture for both the current study and the temporal prior-aware extension is illustrated in Figure~\ref{fig:architecture}.

\begin{figure}[thb]
    \centering
    \begin{adjustbox}{width=\textwidth}
    \begin{tikzpicture}[
        node distance=0.8cm and 0.4cm,
        box/.style={draw, rectangle, rounded corners, minimum width=2.4cm, minimum height=1.0cm, align=center, fill=blue!5!white, font=\small},
        encoder/.style={draw, rectangle, rounded corners, minimum width=2.4cm, minimum height=1.0cm, align=center, fill=orange!10!white, font=\small},
        fuse/.style={draw, circle, minimum size=0.9cm, fill=green!10!white, font=\small},
        arrow/.style={-{Latex[scale=0.8]}, thick},
        group/.style={draw, dashed, inner sep=0.4cm, rounded corners}
    ]
        % Current Study Nodes
        \node (cur_cxr) [box] {Current CXR\\(up to 4 views)};
        \node (cur_clin) [box, right=of cur_cxr] {Current Clinical\\Indication};
        \node (cur_vit) [box, right=of cur_clin] {Current Vitals\\(Numeric)};

        \node (img_enc) [encoder, below=of cur_cxr] {ConvNeXt-V2-Nano\\+ Conv2D Proj};
        \node (text_enc) [encoder, below=of cur_clin] {CXR-BERT\\(CLS Token)};
        \node (vit_enc) [encoder, below=of cur_vit] {Vitals\\Projector};

        \draw [arrow] (cur_cxr) -- (img_enc);
        \draw [arrow] (cur_clin) -- (text_enc);
        \draw [arrow] (cur_vit) -- (vit_enc);

        % Prior Study Nodes (optional temporal branch)
        \node (prv_lbl) [box, right=of cur_vit, xshift=1.2cm] {Prior Labels\\(26-dim vector)};
        \node (prv_cxr) [box, right=of prv_lbl] {Prior CXR\\(up to 4 views)};
        \node (prv_txt) [box, right=of prv_cxr] {Prior Clinical\\Text / Report};
        \node (prv_vit) [box, right=of prv_txt] {Prior Vitals\\(Numeric)};

        \node (lbl_enc) [encoder, below=of prv_lbl] {Linear\\Projection};
        \node (prv_img_enc) [encoder, below=of prv_cxr] {ConvNeXt-V2-Nano\\+ Conv2D Proj};
        \node (prv_text_enc) [encoder, below=of prv_txt] {CXR-BERT\\(CLS Token)};
        \node (prv_vit_enc) [encoder, below=of prv_vit] {Vitals\\Projector};

        \draw [arrow] (prv_lbl) -- (lbl_enc);
        \draw [arrow] (prv_cxr) -- (prv_img_enc);
        \draw [arrow] (prv_txt) -- (prv_text_enc);
        \draw [arrow] (prv_vit) -- (prv_vit_enc);

        % Segment Embeddings and Time Delta
        \node (seg_emb) [box, below=of text_enc, yshift=-0.5cm, fill=purple!10!white] {Segment Embeddings\\(Slots 0--13)};
        \node (time_delta) [box, below=of prv_text_enc, yshift=-0.5cm, fill=yellow!10!white] {Time-Delta Embedding\\(8 buckets)};

        % Sum operations/adding embeddings
        \node (sum_cur) [fuse, below=of img_enc, yshift=-1.5cm] {\Large +};
        \node (sum_prv) [fuse, below=of prv_img_enc, yshift=-1.5cm] {\Large +};

        % Connections to addition nodes
        \draw [arrow] (img_enc) -- (sum_cur);
        \draw [arrow] (text_enc) -- (sum_cur);
        \draw [arrow] (vit_enc) -- (sum_cur);
        \draw [arrow] (seg_emb) -- (sum_cur);

        \draw [arrow] (lbl_enc) -- (sum_prv);
        \draw [arrow] (prv_img_enc) -- (sum_prv);
        \draw [arrow] (prv_text_enc) -- (sum_prv);
        \draw [arrow] (prv_vit_enc) -- (sum_prv);
        \draw [arrow] (time_delta) -- (sum_prv);
        \draw [arrow] (seg_emb) -- (sum_prv);

        % Concatenate / Sequence Node
        \node (concat) [box, below=of sum_cur, xshift=5.2cm, minimum width=9cm, fill=teal!10!white] {Concatenated Token Sequence (V3 Slim / V2 Expanded Layout)};
        \draw [arrow] (sum_cur) -- (concat);
        \draw [arrow] (sum_prv) -- (concat);

        % Transformer Encoder
        \node (trans_enc) [encoder, below=of concat, minimum width=6cm] {2-Layer Transformer Encoder\\(Self-Attention with Key Padding Masking)};
        \draw [arrow] (concat) -- (trans_enc);

        % Decoder & Output
        \node (mldec) [encoder, below=of trans_enc, fill=red!5!white] {MLDecoder Head\\(Group-FC Query-Attention)};
        \node (out) [box, below=of mldec, fill=green!5!white] {26 Multi-label disease logits};

        \draw [arrow] (trans_enc) -- (mldec);
        \draw [arrow] (mldec) -- (out);

        % Grouping
        \node [group, fit=(cur_cxr) (cur_vit) (vit_enc), label=above:\textbf{Current Study Modalities}] {};
        \node [group, fit=(prv_lbl) (prv_vit) (prv_vit_enc), label=above:\textbf{Prior Study Modalities}] {};

    \end{tikzpicture}
    \end{adjustbox}
    \caption{Overview of the CaMCheX prior-aware multi-modal fusion architecture. Encoders extract features from current and historical (prior) modalities. Segment and time-delta embeddings are added, and the tokens are fused via self-attention and decoded by MLDecoder for multi-label classification.}
    \label{fig:architecture}
\end{figure}

\subsection{Problem Formulation}
Let $\mathcal{D} = \{(\mathbf{X}_{img}^{(i)}, \mathbf{T}_{clin}^{(i)}, \mathbf{V}^{(i)}, \mathbf{M}^{(i)}, \mathbf{y}^{(i)})\}_{i=1}^N$ be a dataset of $N$ patient admissions. For each sample:
\begin{itemize}
    \item $\mathbf{X}_{img} \in \mathbb{R}^{S \times 3 \times H \times W}$ is a sequence of chest X-ray (CXR) images representing up to $S \le 4$ views (frontal, e.g., AP/PA, or lateral, e.g., LL) obtained during the current clinical study.
    \item $\mathbf{T}_{clin}$ represents the clinical indication/history text gathered from the emergency department (ED) notes.
    \item $\mathbf{V} \in \mathbb{R}^D$ is a vector of $D = 7$ numeric vital signs (temperature, heart rate, respiratory rate, O2 saturation, systolic blood pressure, diastolic blood pressure, and gender).
    \item $\mathbf{M} \in \{0, 1\}^D$ is a binary mask where $M_j = 1$ indicates that the $j$-th vital sign is missing, and $0$ indicates it is present.
    \item $\mathbf{y} \in \{0, 1\}^C$ is the ground-truth multi-label vector for $C = 26$ chest pathological findings.
\end{itemize}
Our objective is to train a neural network $f_\theta$ that maps the multi-modal input tuple to a vector of predicted probabilities $\hat{\mathbf{y}} = f_\theta(\mathbf{X}_{img}, \mathbf{T}_{clin}, \mathbf{V}, \mathbf{M}) \in [0, 1]^C$ minimizing a multi-label classification loss.

\subsection{Modality-Specific Encoders}

\subsubsection{Chest X-Ray Image Encoder}
To encode the sequence of images $\mathbf{X}_{img}$, we utilize a ConvNeXt-V2-Nano backbone. Because different view positions (frontal vs. lateral) exhibit distinct spatial distributions and diagnostic patterns, we employ a view-routing mechanism:
\begin{equation}
    \mathbf{F}_{img, s} = \begin{cases} 
      \text{Encoder}_{frontal}(\mathbf{X}_{img, s}) & \text{if } \text{VP}_s \in \{\text{AP}, \text{PA}, \text{Frontal}\} \\
      \text{Encoder}_{lateral}(\mathbf{X}_{img, s}) & \text{if } \text{VP}_s \in \{\text{Lateral}, \text{LL}\} \\
      \mathbf{0} & \text{otherwise}
   \end{cases}
\end{equation}
where $\text{VP}_s$ is the view position of slot $s$. For active views, the raw backbone generates spatial features $\mathbf{F}_{img, s} \in \mathbb{R}^{C_{back} \times h_{back} \times w_{back}}$. To integrate these into our transformer architecture, a projection layer (implemented as a 2D convolution with kernel size 3 and stride 2) projects the channels to the transformer's latent dimension $d_{model} = 768$ while reducing the spatial resolution to a grid size of $8 \times 8$:
\begin{equation}
    \mathbf{Z}_{img, s} = \text{Conv2D}(\mathbf{F}_{img, s}) \in \mathbb{R}^{d_{model} \times 8 \times 8}
\end{equation}
We add a 2D sinusoidal spatial positional encoding $\mathbf{P}_{2D} \in \mathbb{R}^{d_{model} \times 8 \times 8}$ to preserve spatial structure:
\begin{equation}
    \mathbf{\tilde{Z}}_{img, s} = \mathbf{Z}_{img, s} + \mathbf{P}_{2D}
\end{equation}
Inactive image slots (e.g., studies with fewer than 4 views) are padded with an all-zero tensor $\mathbf{0}$ and a learnable padding token $\mathbf{t}_{pad} \in \mathbb{R}^{d_{model}}$.

\subsubsection{Clinical Indication Text Encoder}
The clinical text $\mathbf{T}_{clin}$ contains critical contextual information, such as the primary symptoms or reasons for the exam. We leverage a specialized clinical text encoder, BiomedVLP-CXR-BERT-specialized. The text is first tokenized and padded to a maximum length of 384 tokens. The encoder maps this sequence to a high-dimensional space, and we extract the CLS token representation:
\begin{equation}
    \mathbf{h}_{clin} = \text{CXR-BERT}(\mathbf{T}_{clin}) \in \mathbb{R}^{d_{model}}
\end{equation}

\subsubsection{Vital Signs Encoder}
To process the vital signs without relying on noisy text translations, we normalize the raw numeric values $\mathbf{V}$ using statistics (mean and standard deviation) computed over the training set:
\begin{equation}
    \tilde{V}_j = \frac{V_j - \mu_j}{\sigma_j}
\end{equation}
For missing values ($M_j = 1$), we fill $\tilde{V}_j = 0$. The vital signs encoder, \texttt{VitalsTokenProjector}, concatenates the normalized values and the missingness indicator mask to form a feature vector $\mathbf{x}_{vit} = [\mathbf{\tilde{V}}; \mathbf{M}] \in \mathbb{R}^{2D}$. This feature vector is mapped to the embedding space via a Multi-Layer Perceptron (MLP):
\begin{equation}
    \mathbf{Z}_{vit} = \text{MLP}(\mathbf{x}_{vit})
\end{equation}
where the MLP consists of two linear layers with a GeLU activation and Dropout.

\subsection{Multimodal Fusion \& Token Grid Layouts}
The outputs of the modality encoders are mapped into a unified sequence of tokens before being fed to a Transformer Encoder. We compare two distinct token layouts, which represent a trade-off between spatial alignment and computational efficiency.

\subsubsection{Spatially Expanded Fusion (CaMCheX V2)}
In the V2 architecture, we align the text and vitals tokens with the spatial grid of the image features. Since each image view produces a grid of $8 \times 8 = 64$ spatial tokens, the clinical text representation $\mathbf{h}_{clin}$ and the vital signs projection $\mathbf{Z}_{vit}$ are expanded (duplicated) across an identical $8 \times 8$ grid:
\begin{align}
    \mathbf{Z'}_{clin} &= \text{Expand}(\mathbf{h}_{clin}) \in \mathbb{R}^{64 \times d_{model}} \\
    \mathbf{Z'}_{vit} &= \text{MLP}_{v2}(\mathbf{x}_{vit}) \in \mathbb{R}^{64 \times d_{model}}
\end{align}
This strategy produces a long token sequence. For a patient with 4 image views, the sequence contains $4 \times 64 = 256$ image tokens, 64 clinical text tokens, and 64 vital signs tokens, resulting in a total sequence length of 384 tokens. While this enables pixel-to-token cross-attention alignment, it incurs significant quadratic self-attention cost.

\subsubsection{Slim Single-Token Fusion (CaMCheX V3)}
To address the computational bottleneck and evaluate whether spatial expansion carries diagnostic signals, the V3 architecture collapses each non-image modality into a single token:
\begin{align}
    \mathbf{Z'}_{clin} &= \mathbf{h}_{clin} \in \mathbb{R}^{1 \times d_{model}} \\
    \mathbf{Z'}_{vit} &= \text{MLP}_{v3}(\mathbf{x}_{vit}) \in \mathbb{R}^{1 \times d_{model}}
\end{align}
For 4 image views, the sequence length drops from 384 to $256 + 1 + 1 = 258$ tokens. This represents a 33\% reduction in token length, drastically accelerating training and inference.

\subsubsection{Segment Embeddings \& Transformer Fusion}
To distinguish between tokens originating from different modalities and image slots, we add learnable segment embeddings $\mathbf{S} \in \mathbb{R}^{6 \times d_{model}}$ to the sequence:
\begin{equation}
    \mathbf{e}_k = \mathbf{z}_k + \mathbf{s}_{\text{seg}(k)}
\end{equation}
where $\text{seg}(k) \in \{0, 1, 2, 3\}$ corresponds to image view slots, $4$ denotes the clinical indication token, and $5$ represents the vital signs token.
The combined token sequence is fed to a 2-layer Transformer Encoder with 8 attention heads. An attention key padding mask is applied to zero out attention weights for padded image views and missing vital signs:
\begin{equation}
    \mathbf{H} = \text{TransformerEncoder}(\mathbf{E}, \text{Mask})
\end{equation}

\subsection{Prior-Aware Temporal Modeling Extension}
Clinical diagnosis is inherently comparative; clinicians regularly review historical (prior) chest X-rays and clinical reports to determine disease progression. To emulate this clinical workflow, we extend CaMCheX to a prior-aware architecture.

\subsubsection{Inputs \& Information Leakage Prevention}
For patients with a previous study, we retrieve the prior images $\mathbf{X}_{prv}$, prior clinical history $\mathbf{T}_{prv}$, and prior vital signs $\mathbf{V}_{prv}$. Crucially, we also incorporate the prior radiology report $\mathbf{T}_{report}$ and the prior label vector $\mathbf{y}_{prv} \in \mathbb{R}^C$ derived from the prior report. Although the current report cannot be used during current classification as it contains ground-truth labels (representing label leakage), the prior study's report and labels represent legitimate historical context.

\subsubsection{Prior Token Layout and Embedding Projection}
We project the prior multi-label vector $\mathbf{y}_{prv}$ into the embedding space using a linear layer:
\begin{equation}
    \mathbf{Z}_{prv\_label} = \mathbf{W}_{label} \mathbf{y}_{prv} \in \mathbb{R}^{d_{model}}
\end{equation}
The prior report $\mathbf{T}_{report}$ is encoded using the same text encoder to obtain a single CLS token $\mathbf{h}_{prv\_report} \in \mathbb{R}^{d_{model}}$. In the V3 configuration, the complete sequence of tokens for a sample with a prior exam is structured as:
\begin{equation}
    \mathbf{E} = [\mathbf{E}_{cur\_img}, \mathbf{e}_{cur\_clin}, \mathbf{e}_{cur\_vit}, \mathbf{E}_{prv\_img}, \mathbf{e}_{prv\_clin}, \mathbf{e}_{prv\_vit}, \mathbf{e}_{prv\_report}, \mathbf{e}_{prv\_label}]
\end{equation}
Segment embeddings are expanded to 14 slots: 0--3 for current image views, 4--5 for current clinical/vitals, 6--9 for prior image views, and 10--13 for prior clinical, vitals, labels, and report tokens respectively.

\subsubsection{Time-Delta Bucketing}
The diagnostic relevance of a prior study decays as the time gap $\Delta t$ (in days) increases. To allow the transformer to incorporate temporal decay, we discretize $\Delta t$ into 8 buckets:
\begin{equation}
    \text{Bucket}(\Delta t) = \begin{cases}
        0 & \text{if prior is absent / unknown} \\
        1 & \text{if } \Delta t \le 1 \\
        2 & \text{if } 1 < \Delta t \le 7 \\
        3 & \text{if } 7 < \Delta t \le 30 \\
        4 & \text{if } 30 < \Delta t \le 180 \\
        5 & \text{if } 180 < \Delta t \le 365 \\
        6 & \text{if } 365 < \Delta t \le 1095 \\
        7 & \text{if } \Delta t > 1095
    \end{cases}
\end{equation}
A learned time-delta embedding vector $\mathbf{E}_{\Delta} \in \mathbb{R}^{d_{model}}$ is retrieved based on the bucket and added directly to every token in the prior study block.

\subsubsection{Stochastic Prior Dropout}
To ensure that the prior-aware model remains robust when patients have no prior clinical history, we apply stochastic prior dropout during training. With a probability $p_{dropout} = 0.2$, the entire prior block is masked out by setting the corresponding key padding mask values to True, forcing the model to rely solely on the current study features.

\subsubsection{Asymmetric Prior Cross-Attention (CaMCheX V4)}
While the CaMCheX V3 prior-aware model performs full self-attention over the concatenated current and prior sequence $\mathbf{E} \in \mathbb{R}^{518 \times d_{model}}$, this incurs a quadratic computational cost of $\mathcal{O}(518^2)$ attention operations per layer. Furthermore, direct self-attention allows the classification head to easily exploit the prior labels $\mathbf{y}_{prv}$ as a shortcut feature, which poses an overfitting risk (the prior-label copy shortcut).

To alleviate these limitations, we introduce the \textbf{CaMCheX V4} architecture. Instead of concatenating the sequences, V4 treats the current and prior modalities asymmetrically, framing the fusion as a cross-attention decoding task:
\begin{itemize}
    \item \textbf{Target Stream (Queries)}: The current study tokens act as the target sequence $\mathbf{E}_{cur}$ that forms the residual stream:
    \begin{equation}
        \mathbf{E}_{cur} = [\mathbf{E}_{cur\_img}, \mathbf{e}_{cur\_clin}, \mathbf{e}_{cur\_vit}] \in \mathbb{R}^{258 \times d_{model}}
    \end{equation}
    \item \textbf{Memory Stream (Keys/Values)}: The prior study tokens act as the memory sequence $\mathbf{E}_{prv}$, which is never updated during attention layers:
    \begin{equation}
        \mathbf{E}_{prv} = [\mathbf{e}_{sentinel}, \mathbf{E}_{prv\_img}, \mathbf{e}_{prv\_clin}, \mathbf{e}_{prv\_report}, \mathbf{e}_{prv\_vit}, \mathbf{e}_{prv\_label}] \in \mathbb{R}^{261 \times d_{model}}
    \end{equation}
    where $\mathbf{e}_{sentinel} \in \mathbb{R}^{1 \times d_{model}}$ is a learned \texttt{no\_prior\_token} sentinel. This sentinel guarantees at least one valid key/value token when a patient has no prior history (or when prior dropout is active), preventing NaN values in cross-attention, while serving as a learned representation for the absence of prior clinical data.
\end{itemize}

By partitioning the sequence, each fusion layer is structured as a Transformer Decoder layer performing self-attention on the target stream, followed by cross-attention from the target to the prior memory stream:
\begin{align}
    \mathbf{H}_{cur}^{(l, \text{self})} &= \text{SelfAttention}(\text{LN}(\mathbf{H}_{cur}^{(l-1)})) + \mathbf{H}_{cur}^{(l-1)} \\
    \mathbf{H}_{cur}^{(l, \text{cross})} &= \text{CrossAttention}(\text{LN}(\mathbf{H}_{cur}^{(l, \text{self})}), \text{LN}(\mathbf{E}_{prv})) + \mathbf{H}_{cur}^{(l, \text{self})} \\
    \mathbf{H}_{cur}^{(l)} &= \text{FFN}(\text{LN}(\mathbf{H}_{cur}^{(l, \text{cross})})) + \mathbf{H}_{cur}^{(l, \text{cross})}
\end{align}
where $l \in \{1, \dots, L\}$ denotes the layer index and $\text{LN}$ represents LayerNorm (with pre-LN configuration). To prevent the pre-LN outputs from bypassing normalization at the final layer, an explicit final LayerNorm is applied to the output of the decoder stack:
\begin{equation}
    \mathbf{H} = \text{LayerNorm}(\mathbf{H}_{cur}^{(L)}) \in \mathbb{R}^{258 \times d_{model}}
\end{equation}
This asymmetric layout reduces the attention cost from $518^2$ ($\sim$268k operations) to $258^2 + 258 \times 261$ ($\sim$134k operations) per layer—a reduction of approximately 50\%—while preserving every prior spatial patch at its full $8 \times 8$ resolution.

\paragraph{Per-Modality Pre-Fusion Normalization}
Because the fused token stream mixes features from wildly different sources (ConvNeXt-V2 visual features, CXR-BERT text embeddings, numeric vitals, linear label projections, and segment/time-delta embeddings), their numerical scales differ significantly. In a pre-LN Transformer Decoder, the cross-attention keys and values ($\mathbf{E}_{prv}$) do not pass through the main target residual stream's LayerNorm. To resolve this, V4 applies an independent LayerNorm to each modality block to map them to a harmonized unit scale before entering the fusion decoder:
\begin{equation}
    \mathbf{E}'_{type} = \text{LayerNorm}_{type}(\mathbf{E}_{type})
\end{equation}
where $type \in \{\text{img}, \text{clin}, \text{vitals}, \text{report}, \text{label}, \text{sentinel}\}$.

\subsection{Classification Head \& Loss Functions}

\subsubsection{Multi-Label Decoder (MLDecoder)}
Instead of standard pooling followed by a linear layer, we use MLDecoder to decode the transformer representation. MLDecoder constructs a set of learnable query embeddings $\mathbf{Q} \in \mathbb{R}^{K \times d_{model}}$ (where $K < C$ is the number of groups, typically 100). These queries undergo cross-attention with the final transformer tokens:
\begin{equation}
    \mathbf{H}_{dec} = \text{CrossAttention}(\mathbf{Q}, \mathbf{H})
\end{equation}
A group fully-connected layer maps the queried features to the 26 target logits, reducing computation and parameter count compared to traditional query-based decoders like query-attention.

\subsubsection{Asymmetric Loss (ASL)}
To combat severe class imbalance across the 26 labels, we optimize the model using Asymmetric Loss (ASL). ASL dynamically down-weights easy negative examples by applying asymmetric focusing parameters $\gamma_{pos}$ and $\gamma_{neg}$, and shifts negative probabilities with a margin $m$:
\begin{equation}
    L_{ASL} = -\frac{1}{C}\sum_{c=1}^C \left[ y_c w_c^{pos} (1-p_c)^{\gamma_{pos}}\log(p_c) + (1-y_c)w_c^{neg}(p_c^{m})^{\gamma_{neg}}\log(1-p_c^{m}) \right]
\end{equation}
where $p_c = \sigma(\hat{y}_c)$, $p_c^m = \text{clamp}(1 - p_c + m, 0, 1)$, and $w_c^{pos}, w_c^{neg}$ are inverse frequency weights:
\begin{align}
    w_c^{pos} &= \exp(1 - q_c) \\
    w_c^{neg} &= \exp(q_c)
\end{align}
where $q_c$ is the proportion of positive instances for class $c$.
