\section{Discussion}
\label{sec:discussion}

\subsection{Interpretation of Results and Clinical Relevance}
\label{subsec:discussion_interpretation}

We discuss what the results imply clinically and technically, then outline the main limitations and future work.

The quantitative results show that adding clinical text and vital signs improves chest X-ray classification over image-only prediction. The largest improvement comes from adding prior-study context, especially for medium and tail labels. This supports the clinical intuition that previous examinations provide useful comparative evidence for interpreting current radiographs.

From a clinical standpoint, these findings reflect actual diagnostic workflows. Physicians rarely interpret radiographs in isolation; they evaluate findings in light of the patient's presenting symptoms, physiological state, and past imaging history \cite{sloan2025camchex,bannur2023biovilt}. Our asymmetric target-memory fusion lets the current study use prior context as support, which improves pathology prediction.

For tail classes, which suffer from a lack of visual training examples, the clinical text and historical modalities act as regularizers. A rare condition like a hiatal hernia might be misclassified as a lung mass by an image-only model. However, when the model is provided with clinical text explicitly stating ``evaluate for hiatal hernia'' or prior studies showing a stable hernia, it can make a correct prediction. This suggests that multimodal fusion and prior context can reduce, but not solve, the effect of class imbalance.

The temporal label analysis in Section~\ref{subsec:eda_prior_gaps} (Figure~\ref{fig:temporal_label_dynamics}) further clarifies why prior context helps: many findings are temporally persistent, so historical information provides a useful baseline. However, common findings such as Lung Opacity, Atelectasis, and Pleural Effusion also show meaningful rates of new appearance and resolution between studies. The model must therefore still rely on current-image evidence for these transitions, which is consistent with the design choice of using prior context as auxiliary memory rather than as a deterministic prediction source.

\subsection{Limitations}
\label{subsec:discussion_limitations}

The study has several limitations.

\textbf{Reduced Image Resolution}: We resized images to $512 \times 512$ pixels. While sufficient for larger findings like cardiomegaly or pleural effusion, this downsampling may lose fine-grained visual details such as subtle pneumothorax lines, micro-nodules, or early interstitial markings.

\textbf{Frozen Clinical Text Encoder}: The language model (\texttt{BiomedVLP-CXR-BERT}) was kept frozen, or its embeddings were precomputed, to save compute. Consequently, the clinical text representation was not fine-tuned end-to-end for the classification objective, potentially limiting its alignment with the multimodal fusion layers.

\textbf{Label Noise}: The CXR-LT-2023 labels are derived automatically from free-text radiology reports using natural language processing toolkits \cite{holste2024cxrlt}. These automated labels contain inherent noise and occasional misclassifications, which can introduce bias into model training and evaluation.

\textbf{Single-Dataset Validation}: Models were trained and evaluated on splits derived from the MIMIC-CXR dataset. While patient-level splits prevent data leakage, the lack of external validation on datasets from other hospital systems limits the ability to evaluate generalizability to different patient populations and imaging protocols.

\textbf{Shortcut Learning Risk}: While prior-study information improves classification performance, it carries a risk of shortcut learning. The model might learn to rely excessively on historical labels or reports rather than analyzing the current radiograph, which could lead to errors if an acute, new pathology develops that was not present in the past. This was mitigated using prior dropout, which reduces but does not eliminate shortcut risk.

\textbf{Qualitative Interpretability}: Grad-CAM \cite{selvaraju2017gradcam} shows which image regions influence the prediction, but it does not prove clinical reasoning or causality.

\textbf{Limited Training Runs}: The study is also limited by the number of full training runs. Although the selected final model was trained successfully, the available compute did not allow exhaustive factorial ablations or repeated-seed experiments for every architectural choice. As a result, some incremental comparisons indicate development trends rather than isolated causal effects.

\subsection{Future Work}
\label{subsec:discussion_future}

Future research should focus on:

\textbf{Full-Scale Training}: With more compute, training at larger scale with higher-resolution inputs would allow the model to capture fine-grained visual findings and benefit from stronger data augmentation.

\textbf{End-to-End Multimodal Fine-Tuning}: Fine-tuning the clinical text and image encoders jointly would allow tighter semantic alignment between the modalities.

\textbf{High-Resolution and Localized Attention}: Implementing localized attention or patch-based processing would allow the model to process high-resolution radiographs without memory overflow, preserving tiny pathological details.

\textbf{Advanced Temporal Architectures}: Replacing simple cross-attention with recurrent or state-space models (like Mamba \cite{gu2024mamba}) could allow the network to fuse sequences of multiple prior studies rather than a single previous check.

\textbf{Multi-Center Validation}: Evaluating the prior-aware model on external datasets (such as CheXpert or PadChest \cite{bustos2020padchest}) is crucial to verify generalizability across different clinical settings.
