\section{Related Work}
\label{sec:related_work}

This section reviews research areas that are directly relevant to this thesis. First, it discusses single-modal chest X-ray classification, which forms the foundation for automated radiographic interpretation. Second, it reviews multimodal clinical fusion, especially methods that combine radiographs with clinical text or structured patient information. Third, it discusses temporal and prior-aware modelling, which is important because radiologists often compare current studies with previous examinations. Finally, it reviews methods for mitigating class imbalance in long-tailed multi-label medical datasets and summarizes representative CXR-LT 2023 challenge solutions.

\subsection{Single-Modal Chest X-Ray Classification}
\label{subsec:related_single_modal}

Deep learning has been widely studied for automated chest X-ray classification. Early large-scale work was enabled by public datasets such as the NIH ChestX-ray8/ChestX-ray14 dataset, CheXpert, and MIMIC-CXR, which provide chest radiographs with labels extracted from radiology reports \cite{wang2017chestxray8,irvin2019chexpert,johnson2019mimiccxr}. These datasets made it possible to train convolutional neural networks for multi-label prediction of common thoracic findings such as atelectasis, cardiomegaly, consolidation, edema, pleural effusion, and pneumothorax.

Many chest X-ray classification systems are based on convolutional neural networks. Architectures such as ResNet and DenseNet have been commonly used because they can learn hierarchical visual features from medical images while remaining practical to train \cite{he2016resnet,huang2017densenet}. For example, DenseNet-based models have often been used in chest X-ray classification because dense connections encourage feature reuse and improve gradient flow. These models usually take one or more radiographic views as input and produce independent probabilities for each disease label.

More recently, modern visual backbones such as Vision Transformers, Swin Transformers, ConvNeXt, and ConvNeXt-V2 have been explored for medical image analysis. ConvNeXt-V2 is particularly relevant to this project because it combines convolutional design with modern architectural improvements and can provide a strong visual feature extractor while remaining more efficient than very large transformer-based models \cite{woo2023convnextv2}. In this thesis, a lightweight ConvNeXt-V2 variant is used as the image encoder in order to balance visual representation quality and computational feasibility.

Although single-modal chest X-ray classifiers have achieved strong results, they have an important limitation: they treat the radiograph mainly as an isolated image. This does not fully reflect clinical practice, where image interpretation is influenced by clinical indication, symptoms, physiological state, and prior history. Therefore, image-only classification is useful as a baseline, but it may not capture the full diagnostic context available during real chest X-ray interpretation.

\subsection{Multimodal Clinical Fusion}
\label{subsec:related_multimodal}

Multimodal learning aims to combine information from different sources in order to produce richer and more clinically meaningful representations. In radiology, the most common multimodal setting is image-text learning, where chest X-rays are paired with radiology reports. Since radiology reports describe visual findings and clinical impressions, they provide a natural source of supervision for learning image-text representations.

Several works have used contrastive learning to align medical images and text. ConVIRT learns medical visual representations by contrasting paired radiographs and reports, showing that naturally occurring image-report pairs can improve representation learning without requiring additional manual annotations \cite{zhang2022convirt}. GLoRIA extends this idea by learning both global and local image-text representations, aligning image regions with words or phrases from radiology reports \cite{huang2021gloria}. BioViL further improves biomedical vision-language pretraining by using stronger domain-specific text modelling through Chest X-Ray Bidirectional Encoder Representations from Transformers (CXR-BERT) and radiology-specific semantic learning \cite{boecking2022biovil}. MedCLIP relaxes the requirement for strictly paired image-text data by using unpaired medical images and text with a semantic matching objective \cite{wang2022medclip}.

These vision-language approaches show that textual information can improve radiographic representation learning. However, many of them focus on pretraining, retrieval, phrase grounding, or report-based image-text alignment. They do not always directly model the full clinical context available at the time of diagnosis, such as clinical indication text, vital signs, or multi-view study structure. In addition, radiology reports may not always be available during inference because they are usually written after image interpretation. Therefore, using reports as direct input for current-study prediction can create an unrealistic setting if the goal is to support interpretation before the final report is produced.

CaMCheX is the most directly related work to this thesis because it proposes a clinically aligned multimodal chest X-ray classification framework \cite{sloan2025camchex}. Instead of relying only on image-report contrastive learning, CaMCheX combines multi-view chest radiographs with structured clinical information, including clinical indications, patient history, and vital signs. This is closer to the information available in real clinical workflows. This thesis follows the same general motivation, but adapts the approach under limited computational resources by using lightweight image encoders, frozen or precomputed text embeddings, and compact transformer fusion layouts.

\subsection{Temporal and Prior-Aware Medical Modelling}
\label{subsec:related_temporal}

Temporal information is important in radiology because interpretation often depends on comparison with prior studies. A finding on a current chest X-ray may be clinically interpreted as new, stable, improved, or worsened depending on previous examinations. For example, a pleural effusion that is increasing over time may require different clinical attention from one that is unchanged from a prior study. Therefore, prior imaging and historical clinical information can provide useful context for current-study prediction.

Recent biomedical vision-language work has started to incorporate temporal structure. BioViL-T explicitly models prior images and reports when they are available, showing that temporal context can improve downstream tasks such as progression classification, phrase grounding, and report generation \cite{bannur2023biovilt}. The associated MS-CXR-T benchmark further highlights the importance of evaluating whether models understand temporal disease progression, including whether findings are improving, stable, or worsening \cite{bannur2023biovilt}.

Prior-aware modelling can be implemented in different ways. Some approaches compare current and prior images using Siamese encoders or feature-difference representations. Others use attention mechanisms so that current-study features can attend to prior-study features. Cross-attention is especially suitable for this setting because it allows the current examination to act as the target while prior examinations provide contextual memory. This matches the clinical idea that the current image remains the main object of interpretation, while the prior study provides supporting evidence.

However, prior-aware modelling also introduces challenges. First, not every patient has a prior study, so the model must remain robust when historical information is missing. Second, the time gap between current and prior examinations can vary widely, so the model should distinguish between recent and old priors. Third, there is a risk of information leakage if prior labels or reports are used carelessly, although design choices can reduce but not eliminate shortcut risk. For this reason, this thesis treats prior information as auxiliary context and uses time-gap embeddings and stochastic prior dropout to reduce over-dependence on historical information.

\subsection{Mitigating Class Imbalance in Medical Datasets}
\label{subsec:related_class_imbalance}

Chest X-ray classification is naturally a multi-label problem because a single study may contain multiple findings. It is also long-tailed because some abnormalities are common, while others are rare. The CXR-LT challenge highlights this issue by evaluating models on long-tailed multi-label chest X-ray classification, where rare findings and label co-occurrence make the task more difficult than standard balanced classification \cite{holste2024cxrlt,holste2025cxrlt2024}.

A common baseline loss for multi-label classification is binary cross-entropy (BCE), where each label is treated as an independent binary prediction. Although simple and widely used, BCE can be dominated by frequent negative labels. In chest X-ray datasets, most labels are negative for most studies, so the model may learn to focus on easy negatives rather than rare positive findings. This can reduce sensitivity for minority classes, which is problematic in medical settings where rare findings may still be clinically important.

Focal Loss addresses class imbalance by reducing the contribution of easy examples and focusing training on harder examples \cite{lin2017focalloss}. However, standard focal loss was originally designed for dense object detection and does not fully address the asymmetric nature of multi-label classification, where the number of negative labels is usually much larger than the number of positive labels.

Asymmetric Loss (ASL) was proposed specifically for multi-label classification and is therefore highly relevant to this thesis \cite{ridnik2021asymmetric}. ASL applies different focusing behaviour to positive and negative samples, allowing easy negatives to be down-weighted more aggressively while preserving learning from positive labels. This makes it suitable for long-tailed multi-label chest X-ray classification, where the model must learn from rare positive cases without being overwhelmed by the large number of negative labels.

\subsection{CXR-LT 2023 Challenge Solutions}
\label{subsec:related_cxrlt_solutions}

Because this thesis focuses on long-tailed, multi-label chest X-ray classification, reviewing the top methods from the CXR-LT 2023 challenge establishes baseline performance and design choices \cite{holste2024cxrlt}. Built on the MIMIC-CXR database, the challenge evaluated models on 26 findings—including 12 rare classes extracted from clinical reports—using patient-level splits to avoid leakage, with macro Mean Average Precision (mAP) as the primary metric, and Mean Area Under the Receiver Operating Characteristic (mAUROC) as an auxiliary metric \cite{holste2024cxrlt,holste2023cxrlt_physionet}. This section highlights four representative submissions: the top-performing CheXFusion model (which employs a Multi-Label Decoder (ML-Decoder) head) \cite{kim2023chexfusion}, the second-ranked ensemble system \cite{nguyenmau2023advanced}, the robust asymmetric loss approach by Park et al. \cite{park2023robust}, and the text-aligned bag-of-tricks framework by Hong et al. \cite{hong2023bagoftricks}.


\begin{table}
    \centering
    \caption[Representative CXR-LT 2023 challenge solutions]{Representative CXR-LT 2023 challenge solutions. Official aggregate test-phase results and rankings are summarized from the CXR-LT challenge overview \cite{holste2024cxrlt}, while individual method descriptions are cited in each row.}
    \label{tab:cxrlt2023_results}
    \begin{adjustbox}{width=\textwidth}
    \begin{tabular}{clcc}
        \hline
        Rank & Team / Strategy & mAP & mAUROC \\
        \hline
        1 & CheXFusion \cite{kim2023chexfusion} (ConvNeXt-S, multi-view Transformer, ML-Decoder, weighted ASL) & 0.372 & 0.850 \\
        2 & Advanced augmentation and ensemble \cite{nguyenmau2023advanced} (EfficientNetV2/ConvNeXt, test-time augmentation (TTA)) & 0.354 & 0.836 \\
        4 & Robust Asymmetric Loss \cite{park2023robust} (ConvNeXt-B, Robust Asymmetric Loss) & 0.351 & 0.838 \\
        5 & Bag of Tricks \cite{hong2023bagoftricks} (ResNet50, text encoder, external data, TTA/ensemble) & 0.349 & 0.836 \\
        \hline
    \end{tabular}
    \end{adjustbox}
\end{table}

The challenge results also showed a clear gap between common and rare findings. Table~\ref{tab:cxrlt2023_longtail} reports mAP grouped by label frequency for the representative teams. The top-performing model achieved 0.499 mAP on head classes but only 0.242 mAP on tail classes. This large performance gap across all methods confirms that rare-label recognition remains extremely challenging even for the strongest systems.

\begin{table}
    \centering
    \caption[CXR-LT 2023 performance by label-frequency group]{CXR-LT 2023 performance by label-frequency group for representative teams. The grouped mAP values are summarized from the CXR-LT challenge overview \cite{holste2024cxrlt}.}
    \label{tab:cxrlt2023_longtail}
    \begin{tabular}{lcccccc}
        \hline
        Team & Year & Resolution & Architecture & Overall mAP & Head mAP & Medium mAP & Tail mAP \\
        \hline
        CheXFusion \cite{kim2023chexfusion} (Rank 1) & 2023 & ConvNeXt, Transformers & 0.499 & 0.246 & 0.242 \\
        Nguyen-Mau et al. \cite{nguyenmau2023advanced} (Rank 2) & 2023 & 0.482 & 0.226 & 0.227 \\
        Park et al. \cite{park2023robust} (Rank 4) & 2023 & 0.480 & 0.221 & 0.220 \\
        Hong et al. \cite{hong2023bagoftricks} (Rank 5) & 2023 & 0.474 & 0.218 & 0.243 \\
        CaMCheX \cite{sloan2025camchex} (Reference) & 2025 & 0.576 & 0.786 & 0.550 & 0.408 \\
        \hline
    \end{tabular}
\end{table}

Several lessons can be drawn from these challenge systems. Most top-performing models utilized ConvNeXt or Convolutional Neural Network (CNN)-family backbones, and their performance was heavily boosted by high image resolutions, test-time augmentations (TTA), and model ensembling \cite{holste2024cxrlt}. However, even the highest-ranked models showed weak performance on rare classes, confirming that tail-class classification remains an unresolved challenge. Rather than relying on leaderboard-style ensembling or high-resolution images, this thesis instead studies a resource-constrained multimodal approach, investigating if clinical indications, vital signs, and prior-study context can improve prediction accuracy.

\subsection{CaMCheX Results and Relevance to This Thesis}
\label{subsec:related_camchex_results}

CaMCheX is the primary reference for this thesis because it combines multi-view chest radiographs with clinical indications and vital signs using a multimodal transformer-based framework \cite{sloan2025camchex}. As summarized in Table~\ref{tab:camchex_reference_results}, CaMCheX reported 0.576 mAP and high Area Under the Receiver Operating Characteristic (AUROC) values on CXR-LT 2023 and 0.725 mAP on CXR-LT 2024, demonstrating that multimodal context can substantially improve classification accuracy. These benchmark results are included as reference context, but they are not directly comparable to the experiments in this thesis due to differences in dataset scale and training resources.

\begin{table}
    \centering
    \caption[Reference results reported by CaMCheX on classification benchmarks]{Reference results reported by CaMCheX on chest X-ray classification benchmarks. These numbers are included as prior-work context and are not directly comparable to the resource-constrained experiments in this thesis.}
    \label{tab:camchex_reference_results}
    \begin{tabular}{lccc}
        \hline
        Benchmark & Model & mAP & AUROC \\
        \hline
        CXR-LT 2023 & CheXFusion & 0.372 & 0.850 \\
        CXR-LT 2023 & CaMCheX & 0.576 & 0.916 \\
        CXR-LT 2024 Task 2 & Best CXR-LT 2024 competitor & 0.526 & 0.833 \\
        CXR-LT 2024 Task 2 & CaMCheX & 0.725 & 0.904 \\
        MIMIC-CXR labels & CaMCheX & 0.669 & 0.934 \\
        \hline
    \end{tabular}
\end{table}

As shown in Table~\ref{tab:camchex_longtail_results}, the performance of CaMCheX varies across frequency groups, achieving high mAP on head classes but lower mAP on body and tail classes. This pattern confirms that long-tailed classification remains difficult even for strong multimodal models.

\begin{table}
    \centering
    \caption{CXR-LT 2023 frequency-group mAP reported by CaMCheX.}
    \label{tab:camchex_longtail_results}
    \begin{tabular}{lcccc}
        \hline
        Model & Overall mAP & Head mAP & Body mAP & Tail mAP \\
        \hline
        CheXFusion & 0.372 & 0.499 & 0.246 & 0.242 \\
        CaMCheX & 0.576 & 0.786 & 0.550 & 0.408 \\
        \hline
    \end{tabular}
\end{table}

While these results motivate the design choices adopted in this thesis---including multi-view encoders, transformer-based fusion, and asymmetric loss---this work does not attempt an exact reproduction. The original CaMCheX framework relies on large-scale training with high-resolution inputs and external pretraining. This thesis instead implements and evaluates a resource-constrained adaptation, extending the architecture with prior-study temporal context.
