\section{Exploratory Data Analysis}
\label{sec:eda}

This section details the properties of the datasets utilized in this study, the cohort selection criteria, and statistical analyses of the image, text, vital signs, and temporal features.

\subsection{Dataset Cohort and Subsets}
\label{subsec:eda_cohort}

Describe the source datasets: MIMIC-CXR-JPG (radiographs and reports) and MIMIC-IV / MIMIC-IV ED (demographics, vital signs, and clinical indications). Explain the cohort selection criteria: matching patient admissions that contain concurrent CXR studies, ED admission vitals, and clinical history text.
\begin{itemize}
    \item Detail the subset extraction process (fraction, sampling strategy, and size of train/val/test splits).
    \item Include a summary table showing cohort counts, total studies, and missingness rates of different modalities.
\end{itemize}

% [TABLE PLACEHOLDER]
% Table name: tab:cohort_summary
% Description: Overall cohort statistics (number of patients, studies, views, and text/vitals availability).

\subsection{Label Distribution and Co-occurrence}
\label{subsec:eda_labels}

Analyze the 26 target pathology findings derived from the CXR-LT (Chest X-ray Long-Tail) annotations:
\begin{itemize}
    \item Describe the distribution of positive labels across the cohort, highlighting the long-tailed nature of the pathologies (e.g., Cardiomegaly, Pleural Effusion vs. rare findings).
    \item Explain the co-occurrence patterns between different findings.
\end{itemize}

% [FIGURE PLACEHOLDER]
% Figure name: fig:label_distribution
% Description: Bar plot showing the frequency (prevalence) of the 26 target pathology labels.

% [FIGURE PLACEHOLDER]
% Figure name: fig:label_correlation
% Description: Heatmap showing the correlation/co-occurrence matrix of the 26 pathology classes.

\subsection{Clinical Indication Text Analysis}
\label{subsec:eda_text}

Analyze the emergency department clinical indication notes:
\begin{itemize}
    \item Discuss text properties: character/word counts, sentence counts, and sequence lengths.
    \item Identify common vocabulary, key clinical terms (e.g., "shortness of breath", "cough", "fever", "chest pain"), and typical syntax/structure.
\end{itemize}

% [FIGURE PLACEHOLDER]
% Figure name: fig:text_length_distribution
% Description: Histogram of word/token count distributions for the clinical indication text.

\subsection{Vital Signs and Missingness Analysis}
\label{subsec:eda_vitals}

Analyze the distributions of the 7 vital signs (temperature, heart rate, O2 saturation, respiratory rate, systolic blood pressure, diastolic blood pressure, and gender):
\begin{itemize}
    \item Present statistics (mean, standard deviation, min, max, median) for each vital sign.
    \item Detail the missingness patterns: which vitals are most frequently missing (e.g., O2 saturation or respiratory rate) and the overall percentage of missing values.
\end{itemize}

% [TABLE PLACEHOLDER]
% Table name: tab:vitals_statistics
% Description: Summary statistics and missingness percentages for each of the 7 vital signs.

% [FIGURE PLACEHOLDER]
% Figure name: fig:vitals_distribution
% Description: Box plots or density plots for each numeric vital sign.

\subsection{Temporal Prior Gaps}
\label{subsec:eda_temporal}

Analyze the temporal dynamics of the dataset for patients with prior studies:
\begin{itemize}
    \item Describe the frequency of prior studies per patient (how many patients have at least one prior study).
    \item Analyze the distribution of the time gap ($\Delta t$ in days) between the current study and the most recent prior study. Explain why discretization/bucketing is appropriate given the skewed distribution of temporal intervals.
\end{itemize}

% [FIGURE PLACEHOLDER]
% Figure name: fig:time_delta_distribution
% Description: Histogram showing the distribution of time gaps (in days) between current and prior studies.
